Efficient Bot for The Game of San Jego\\\large Algorithm Engineering LAB 2021 Project Paper

### 2021-02-28 paper/paper.tex
Update remarks in paper

The five-finger pattern \cite{macgilchrist2014}:
\begin{enumerate}
\item \textbf{Topic and background:} What topic does the paper deal with? What is the point of departure for your research? Why are you studying this now?
\item \textbf{Focus:} What is your research question? What are you studying precisely?
\item \textbf{Method:} What did you do?
\item \textbf{Key findings:} What did you discover?
\item \textbf{Conclusions or implications:} What do these findings mean? What broader issues do they speak to?
\end{enumerate}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{combinatorial game theory, optimization, bot}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

@@

\subsection{Related Work}
San Jego must be viewed in the broader context of board game research.
Engines for these games typically perform three kinds of actions: move generation, search and evaluation \cite{Bimonugroho2020}.
% more on that

For instance in chess, relevant features of the board are commonly represented as bitboards where each field corresponds to one bit in a 64 bit integer \cite{Bimonugroho2020}.
One such bitboard may track the fields that have black pieces on them, and another one may track the positions of knights.
Finding all black knights on the board is now as efficient as computing a bitwise AND of these two bitboards.

This particular game has not received much attention yet.
Major contributions were made by \citeauthor{Althöfer2020} \cite{Althöfer2020}.

@@

\section{Theoretical Analysis}
\paragraph{Tower design}
This first paragraph focuses on the central objects of this game: the towers.
Towers are built from the bricks the game starts with by stacking them onto another.
Important operations on them include the retrieval of height and color of the top brick, as the latter identifies the owner of the tower, and obviously an operations to place them on top of each other.
Moreover, as we will see later, it is beneficial to offer a way to \enquote{subtract} towers from each other, that is revert the stacking.

The naive implementation would use a growable, list-like structure of integers to store the bricks.
This would allow constant access times to height and top brick, and linear time complexity to merge two towers.
As the bricks can only be in one of two states, namely blue and yellow, this would waste a lot of memory.

@@

Although the number of moves in a game is usually much lower, this can possibly be used for time management to estimate the remaining length of a game.

@@

Towers that have no neighbours can easily be ignored during move generation as they will not be able to move ever again.

@@

This behaviour is well known from the related game Clobber \cite{Althöfer2004}.

@@

@@

\begin{itemize}
  %\item board is symmetrical and rotationally invariant -> speed up start
  %\item towers can not move to empty fields -> each turn a tower is removed -> max number of turns is fixed and depends on board size
  \item what move to choose in first turn? all moves produce same board value -> empirically: moves to center of boards are more powerful on average \cite{Althöfer2020}
  \item maximum tower height: m*n => can be used for representation improvement
  %\item only an idea, so better rethink: cache the hights of the highest towers of both players (avoids costly recalculation in every explored node); recalculation only necessary if new higher tower was built or a tower with the max height of a player was taken
  %\item if a player had to skip once, there will never be a move available to them again
  %\item isolated regions
\end{itemize}

@@

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{literature}

\end{document}
\endinput
%%
%% End of file `sample-sigconf.tex'.

>>>>

\begin{enumerate}
\item \textbf{Topic and background:} What topic does the paper deal with? What is the point of departure for your research? Why are you studying this now?
\item \textbf{Focus:} What is your research question? What are you studying precisely?
\item \textbf{Method:} What did you do?
\item \textbf{Key findings:} What did you discover?
\item \textbf{Conclusions or implications:} What do these findings mean? What broader issues do they speak to?
\end{enumerate}

@@

\keywords{combinatorial game theory, optimization, bot}

\maketitle

@@

\subsection{Related Work}
San Jego must be viewed in the broader context of board game research.
Engines for these games typically perform three kinds of actions: move generation, search and evaluation \cite{Bimonugroho2020}.
% TODO more on that

For instance in chess, relevant features of the board are commonly represented as bitboards where each field corresponds to one bit in a 64 bit integer \cite{Bimonugroho2020}.
One such bitboard may track the fields that have black pieces on them, and another one may track the positions of knights.
Finding all black knights on the board is now as efficient as computing a bitwise \texttt{AND} of these two bitboards.

This particular game has not received much attention yet.
Major contributions were made by \citeauthor{Althöfer2020} \cite{Althöfer2020}.
They proved an upper bound for the game state complexity, the branching behavior, and precise values for games on small boards.
The optimal move lines that lead to these values were not published, though.

@@

\section{Theoretical Analysis}
This section is subdivided into two parts.
First, some low-level features of the game are discussed, and it is investigated how they can be used to increase the efficiency of San Jego search programs.
After that, it is shown how these individual parts fit into the greater picture of well-known algorithms for combinatorial games.

\subsection{Low-level considerations}\label{subsec:low-level}
\paragraph{Tower design}
This first paragraph focuses on the central objects of this game: the towers.
Towers are built from the bricks the game starts with by stacking them onto another, meaning that the maximum height they can reach is bounded by the board size.
Important operations on them include the retrieval of height and color of the top brick, as the latter identifies the owner of the tower, and obviously an operations to place them on top of each other.
Moreover, as we will see later, it is beneficial to offer a way to \enquote{subtract} towers from each other, that is revert the stacking.

The naive implementation would use a growable, list-like structure of integers to store the bricks.
This would allow constant access times to height and top brick, and linear time complexity to merge two towers.
As the bricks can only be in one of two states, namely blue and yellow, this would waste a lot of memory.
The next logical step is to store only a single bit for each brick using an appropriate integer type.
It is not practical, however, to just assign 0 and 1 to the available colors, as leading zeros in an integer would cause a loss of information.
Hence, the mapping must be normalized so that the top brick is always represented by a 1, and the actual color must be encoded separately.
Using this representation, all operations can be carried out in constant time using bit manipulations such as the usual \texttt{OR}, shifts and (masked) \texttt{XOR} as well as the advanced \texttt{LZCNT} (leading zeros count) which are mapped to efficient instructions on a broad variety of architectures.
A last modification is possible if the actual structure of bricks in a tower does not matter as it is the case for the standard ruleset considered in this paper.
Then, the integer can be used to store the height value directly, reducing the space complexity logarithmically.
Adding two towers now simplifies to adding their heights and setting the top brick correctly.

@@

Although the number of moves in a game is usually somewhat lower (on average 72\% of the maximum, although the number is higher if both players are good \cite{Althöfer2020}) this can possibly be used for time management to estimate the remaining length of a game.

@@

Towers that have no neighbors can easily be ignored during move generation as they will not be able to move ever again.

@@

This behavior is well known from the related game Clobber \cite{Althöfer2004}.

@@

\subsection{High-level considerations}
\paragraph{Search algorithm}
Searching game trees typically comes in two flavors:
\enquote{full} search and Monte Carlo search (or a combination thereof).
The former searches every possible game state that is reachable from the current state, and returns a path that leads to the best outcome for the active player.
It is most effective if the game state complexity is manageable.
Otherwise, this method may not find the best (or even a good) move within acceptable time.
It is required to use this method, if exact game values are needed, e.g. in theoretical analysis of the game.
In practical settings, some form of $\alpha$-$\beta$-pruning is always applied as it cuts off branches of the search tree that will not influence the outcome.
For the pruning to work most effectively, the best move has to be searched first.
This leads to a chicken-and-egg problem that is typically solved by using a heuristic for the move ordering as discussed in the next paragraph.

Monte Carlo tree search (MCTS) is applied in settings where there are too many states to search completely (or even pruned).
Its core idea is to consider random moves, collect statistics about their outcomes, and choose the move that has the highest probability of being good.

In San Jego, the game state space becomes unmanageable quite early.
\citeauthor{Althöfer2020} report that the full search on boards as small as $5\times5$ with their NegaScout implementation (an optimized version of $\alpha$-$\beta$-pruning) took \enquote{several days} \cite[p.~8]{Althöfer2020}.
Using the optimizations outlined in section \ref{subsec:low-level}, it is likely possible to bring that number down.
Typical computer olympiad matches of similar games, however, are played on $10\times10$ boards, which is completely hopeless to encounter with full search.
Therefore, MCTS should be used for San Jego playing programs.
One important property here is that the branching factor monotonically decreases in the course of the game, as each turn a tower is effectively eliminated.
This makes full tree search applicable in the end game to play that phase more precisely.

\paragraph{Search strategy}
An important factor for the success of game tree searching is the evaluation function.
In some combinatorial games like chess and Go, the value of a game state is generally not known.
Other games, as San Jego, define a value for each game state, but it might not be good approach for sorting candidate moves.
For instance, consider the first turn in a game.
Whatever move the player chooses, they all lead to a game state with the same value of $+1$.
Therefore, several improvements are possible.
\citeauthor{Althöfer2020} found that a combination of prioritizing moves that target high enemy towers and deprioritizing moves that target high own towers on the one hand, and prioritizing moves targetting the center of the board on the other hand work together nicely.

@@

\begin{itemize}
  %\item board is symmetrical and rotationally invariant -> speed up start
  %\item towers can not move to empty fields -> each turn a tower is removed -> max number of turns is fixed and depends on board size
  %\item what move to choose in first turn? all moves produce same board value -> empirically: moves to center of boards are more powerful on average \cite{Althöfer2020}
  %\item max tower height: m*n => can be used for representation improvement
  %\item cache the hights of the highest towers of both players (avoids costly recalculation in every explored node); recalculation only necessary if new higher tower was built or a tower with the max height of a player was taken
  %\item if a player had to skip once, there will never be a move available to them again
  %\item isolated regions
\end{itemize}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{literature}

\end{document}
\endinput
